---
title: "4. Analyse Drosben survival data "
author: "Terrence M. Trinca and Joaquín de Navascués"
output:
  html_document:
    #theme: united
    toc: yes
    toc_float: yes
    toc_depth: 3
    highlight: haddock
---

## 4A. Setting up

### Computing tools

If you have not used R before, you are probably reading this in its HTML version
and you do not know how to run an R markdown notebook. If this is the case, our
recommendation is that you first install R and RStudio from [the Comprehensive R
Archive Network, CRAN](https://cran.r-project.org/) and the [Posit Software,
PBC](https://posit.co/download/rstudio-desktop) download sites, respectively,
and follow their instructions. It just takes a couple of minutes. If you feel a
bit lost, the R-ladies' Sydney chapter has a nice ['basic-basic' tutorial to
using RStudio](https://rladiessydney.org/courses/01-basicbasics-0).

Once you have done this, you can just double-click on the 'Rmd' version of this
notebook to run it as you ran the Jupyter notebooks.

All the analysis examples here are just using the work of (mostly)
Dr Terry M Therneau
([Survival package](https://cran.r-project.org/web/packages/survival/))
and (again, mostly) Alboukadel Kassambara
([Survminer package](https://cran.r-project.org/web/packages/survminer/))

```{r setup_installation_loading, echo=FALSE, cache=FALSE, warning=FALSE}
require(librarian)
librarian::shelf(
  tidyverse,
  readxl,
  survival, survminer,
  ggthemes, ggtext, RColorBrewer,
  knitr,
  rstudioapi, # TMT
  gridExtra,  # TMT
  grid,       # TMT
  quiet = TRUE)
ggplot2::theme_set(ggpubr::theme_pubr(base_size=10))
fsep <- .Platform$file.sep
knitr::opts_chunk$set(
  dev = 'png', 
  fig.align = 'center', fig.height = 7, fig.width = 10, 
  pdf.options(encoding = "ISOLatin9.enc"),
  fig.path=paste0('notebook_figs', fsep),
  warning=FALSE,
  message=FALSE)
```

### Survival analysis

We are not statisticians, so the analyses illustrated here are not very
sophisticated. However, you should be somewhat familiar with:
- [Kaplan-Meier](https://rmvpaeme.github.io/KaplanMeier_intro/) [(KM) plots](pmc.ncbi.nlm.nih.gov/articles/PMC5045282/pdf/jadp-07-091.pdf)
- [Cox-proportional hazards (CPH) modelling](http://www.sthda.com/english/wiki/cox-proportional-hazards-model)
- and [CPH model assumptions](http://www.sthda.com/english/wiki/cox-model-assumptions)

There are plenty of resources online that explain at different levels of
expertise what these are and how to interpret them. Two reasonable resources are
- STHDA's [_Survival Analysis Basics_](https://www.sthda.com/english/wiki/survival-analysis-basics)
and [_Survival Analysis_](https://www.sthda.com/english/wiki/survival-analysis)
guides.
- TDV Swinscow's [_Statistics at Square One_ at the BJM](https://www.bmj.com/about-bmj/resources-readers/publications/statistics-square-one)
But, in a nutshell, KM plots are used to estimate and visualise survival distributions. Log-rank tests can tell you whether two or more of these distributions are significantly different, but not by how much nor is suitable for more than pair-wise comparisons. CPH modelling can also test for significance but also it can (1) compare multiple distributions at time and (2) _quantify the size of the effect_. However, this is only informative if the assumptions of the CPH model hold.

Our steps will be:
- Load the data and demonstrate how to do additional cleanup, if needed.
- Check for equal distribution of replicates within each cohort and decide whether outlier replicates should be eliminated.
- Plot KM graphs to visualise survival distributions.
- Perform CPH modelling for replicates, experimental variables and their interactions.
- Check that the assumptions of the CPH model hold.

## 4B. Initial inspection of the data

### Loading the data

The [last (3^rd^) Jupyter python notebook](./03_Read_DataSheets.html) produces an XLSX file whose `Observations_analysis` spreadsheet will contain the events (deaths, censorings) ready for analysis with _survival_ in the form:

 Day_no. | Rack_no. | Tube_no. | genotype | treatment | Event_code
---------|----------|----------|----------|-----------|------------
    1    |     1    |    1     | wild-type|    mock   | 1 _(death)_
    1    |     1    |    2     | wild-type|    mock   | 0 _(censor)_

You will find this file in:
```
your-home-directory/
 └──Drosben_userdata/
    └──experiments_data/
       └──02.02.2022_Effect-of-myfavouritegene-on-lifespan_bd8b2f79/
          └──bd8b2f79_observations.xlsx
```
```{r,  echo = FALSE, message = FALSE, warning = FALSE}
# This code block allows you to find the file interactively
#filepath <- file.choose()
filepath <- file.path(
  "/Users/JQ",
  "Drosben_userdata/experiments_data",
  "02.02.2022_Effect-of-myfavouritegene-on-lifespan_bd8b2f79/bd8b2f79_observations.xlsx"
  )
survdata <- read_excel(filepath, sheet = 'Observations_analysis')
head(survdata)
```
The column names of this dataframe are fixed, except for the variable names, which are decided by the user in the [first Jupyter noteboook](./01_First_Use_Colour_Calibration.html). We could obtain them from the `experimentID_infodict.xlsx` file, but it is easier to do:
```{r}
fixed_colnames <- c("Day_no.", "Rack_no.", "Tube_no.", "Event_code")
var_names <- !(colnames(survdata) %in% fixed_colnames)
vars <- setdiff(colnames(survdata), fixed_colnames)
var1 <- vars[[1]]
var2 <- vars[[2]]
vars
```

In the example, the specimens are organised like this:

Stratum no. | Variable levels   | Rack_no. | Tube_nos. | Replicates
------------|-------------------|----------|-----------|----------------------
     1      |wild-type, mock    |     1    |    1-12   | rack 1, tubes 1 - 12
     2      | _mfg^1^_, mock    |     2    |    1-12   | rack 2, tubes 1 - 12
     3      |wild-type, treated |     3    |    1-12   | rack 3, tubes 1 - 12
     4      | _mfg^1^_, treated |     4    |    1-12   | rack 4, tubes 1 - 12

So, the replicate could be equated to the `Tube_no.`.
However, the user has the flexibility to organise this in any other way, for instance:

Stratum no. | Variable levels   | Rack_no. | Tube_nos. | Replicates
------------|-------------------|----------|-----------|----------------------
     1      |wild-type, mock    |     1    |    1-6    | rack 1, tubes 1 - 6
            |                   |     2    |    1-6    | rack 2, tubes 1 - 6
------------|-------------------|----------|-----------|----------------------
     2      | _mfg^1^_, mock    |     1    |    7-12   | rack 1, tubes 7 - 12
            |                   |     2    |    7-12   | rack 2, tubes 7 - 12
------------|-------------------|----------|-----------|----------------------
     3      |wild-type, treated |     3    |    1-6    | rack 3, tubes 1 - 6
            |                   |     4    |    1-6    | rack 4, tubes 1 - 6
------------|-------------------|----------|-----------|----------------------
     4      | _mfg^1^_, treated |     3    |    7-12   | rack 3, tubes 7 - 12
            |                   |     4    |    7-12   | rack 4, tubes 7 - 12
------------|-------------------|----------|-----------|----------------------

Or even more complicated arrangements, like each Stratum occupying 3 tubes in each of the 4 racks—maybe not even the same tube numbers in each rack. To ensure _any_ arrangement of replicates (the distinct physical tubes) is captured, we do:
```{r,  echo = FALSE, message = FALSE, warning = FALSE}
rep_key <- survdata %>%
  distinct(.data[[var1]], .data[[var2]], `Rack_no.`, `Tube_no.`) %>%
  arrange(.data[[var1]], .data[[var2]], `Rack_no.`, `Tube_no.`) %>%
  group_by(.data[[var1]], .data[[var2]]) %>%
  mutate(replicate = row_number()) %>%
  ungroup()

survdata <- survdata %>%
  left_join(rep_key, by = c(var1, var2, "Rack_no.", "Tube_no.")) %>%
  select(!c(`Rack_no.`, `Tube_no.`)) %>%
  rename(
    event = Event_code,
    day   = `Day_no.`)

head(survdata)
```

### Analysis of replicates

To ensure consistency of behaviour of within replicates, we subset the data:

```{r}
# replicate as a factor → discrete colours + a clean legend
levels <- as.character(sort(unique(as.integer(as.character(survdata$replicate)))))
survdata$replicate <- factor(survdata$replicate, levels = levels)

fit <- survfit(Surv(time = day, event = event) ~ replicate, data = survdata)

g <- ggsurvplot_facet(
  fit, survdata,
  facet.by      = c(var1, var2),   # one panel per stratum (var1/var2 are strings)
  axis.labels   = "all",
  color         = "replicate",     # colour each replicate to spot outliers
  conf.int      = FALSE,           # keep curves clean, like the base-R version
  censor        = TRUE,            # + censoring marks
  break.time.by = 20,              # tick spacing, echoing axis(seq(10,90,20))
  xlab          = "Lifespan (days)",
  ylab          = "Survival probability",
  short.panel.labs = TRUE,         # panel strips show just the level
  ggtheme       = theme_classic()  # minimal, base-R-like: axis lines, no grid
) +
  # to control the order of colours and replicate number
  scale_colour_manual(
    values = brewer.pal(12, "Paired"),
    name   = "Replicate",
    limits = levels,
    breaks = levels
  ) +
  # to plot internal XY axes
  annotate("segment", x = -Inf, xend = -Inf, y = -Inf, yend =  1,
           linewidth = 0.5) +                         # y-axis, left edge
  annotate("segment", x = -Inf, xend =  Inf, y = -Inf, yend = -Inf,
           linewidth = 0.5) +                          # x-axis, bottom edge
  coord_cartesian(clip = "off")

g
```

From this, we might want to remove replicate #2 of the _mfg^1^_, mock stratum, as there is full mortality beforre day 40. However, this is likely due to a large amount of censoring at around day 20, followed by the death of one or two individuals left in the tube (see [Jupyter notebook #3](./03_Read_DataSheets.html), section 3C). This will likely have no impact on the final result, but we might just as well eliminate this replicate:
```{r}
survdata_filt <- survdata %>%
  filter_out(.data[[var1]]=='mfg[1]' & .data[[var2]]=='mock' & replicate==2)

fit <- survfit(Surv(time = day, event = event) ~ replicate, data = survdata_filt)

g <- ggsurvplot_facet(
  fit, survdata_filt,
  facet.by      = c(var1, var2),   # one panel per stratum (var1/var2 are strings)
  axis.labels   = "all",
  color         = "replicate",     # colour each replicate to spot outliers
  conf.int      = FALSE,           # keep curves clean, like the base-R version
  censor        = TRUE,            # + censoring marks
  break.time.by = 20,              # tick spacing, echoing axis(seq(10,90,20))
  xlab          = "Lifespan (days)",
  ylab          = "Survival probability",
  short.panel.labs = TRUE,         # panel strips show just the level
  ggtheme       = theme_classic()  # minimal, base-R-like: axis lines, no grid
) +
  scale_colour_manual(
    values = brewer.pal(12, "Paired"),
    name   = "Replicate",
    limits = levels,
    breaks = levels
  )

g
```
While replicates #1 for both treatments in _mfg^1^_ individuals seem to be outliers, they are not fundamentally different and I will keep them.


## 4C. Effect on survival of `r var1`, `r var2` or their interaction











Prepare the graph:

```{r,  echo = FALSE, message = FALSE, warning = FALSE}
palette <- c('#85C1E9', '#999999', '#E74C3C', #'#F1C40F'
             '#4444DD', '#000000', '#A2103C') # '#F39C12'
line_type <- c(rep("solid",3),rep("twodash",3))
title <- 'Susceptibility to *S. marcescens*'
g <- ggsurvplot(surv_fit(Surv(hour, event) ~ genotype + treatment,
                         data = data),
                pval = TRUE,                       # adds pval
                conf.int = TRUE,                   # adds 95% confidence interval
                conf.int.alpha = 0.1,              # makes it more transparent
                surv.median.line = "v",           # adds median points
                palette = palette,
                xlab = "Time (hours)",              # remove x axis label
                ylab = "Survival Probability",     # y axis label
                linetype = line_type,
                title = title,
                font.legend = c(7),               # legend size
                legend = "bottom")                  # legend position
```

#### To prepare the legend:

Let's have a look at our strata (by default organised as `genotype:treatment` in alphabetical order): 
```
{r,  echo = FALSE, message = FALSE, warning = FALSE}
levels(g$data.survtable$strata)
```
But I'd like to present the strata as here:
```
{r,  echo = FALSE, message = FALSE, warning = FALSE}
exgrid <- expand.grid(unique(sort(data$genotype, decreasing = TRUE)),
                      unique(sort(data$treatment, decreasing = TRUE)))
exgrid <- exgrid[order(exgrid$Var2),]
lab_strata <- paste(exgrid$Var1, exgrid$Var2, sep=" | ")
lab_strata
```
```
{r}
lab_strata <- c("*myoIA>ctrl^KD^* | Sucrose",
                "*myoIA>Etl1^KD^* | Sucrose",
                "*ken^1^* | Sucrose",
                "*myoIA>ctrl^KD^* | S.marcescens",
                "*myoIA>Etl1^KD^* | S.marcescens",
                "*ken^1^* | S.marcescens")
lab_strata
```
to match the colours and the line types like this:

```
{r,  echo = FALSE, message = FALSE, warning = FALSE, fig.height = 1, fig.width=2.5}
x <- rep(c(1:20), 6)
variable <- unlist(lapply(lab_strata, rep, 20))
value <- c( 5+rnorm(20)/5, 4+rnorm(20)/5, 3+rnorm(20)/5,
            2+rnorm(20)/5, 1+rnorm(20)/5, rnorm(20)/5 )
df <- data.frame(x, variable, value)
d <- ggplot(df, aes(x=x, y=value, group=variable, colour=variable )) +
  geom_line(aes(linetype=variable, colour=variable)) +
  scale_linetype_manual(breaks=lab_strata, values=line_type, labels=LETTERS[1:6]) +
  scale_colour_manual(breaks=lab_strata, values=palette, labels=LETTERS[1:6]) +
  theme(legend.text=element_text(size=5),
        legend.title = element_text(size=6),
        legend.key.height = unit(1/6, 'in'))
d
```

To match what we want, the order of the legends, colours and line types should be like:
```
{r,  echo = FALSE, message = FALSE, warning = FALSE}
lab_strata[c(1,3,2,4,6,5)]
```
And the reverse order would be:
```
{r,  echo = FALSE, message = FALSE, warning = FALSE}
l <- levels(g$data.survtable$strata)[c(6,2,4,5,1,3)]
l
```
Now we can start with the preliminary plot:

```
{r,  echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3, fig.width=5}
g$plot +
  scale_colour_manual(name='', breaks=l, values=palette,
                      labels=lab_strata[c(1,3,2,4,6,5)],
                      guide=guide_legend(nrow = 3)) +
  scale_fill_manual(name='', breaks=l, values=palette,
                    labels=lab_strata[c(1,3,2,4,6,5)],
                    guide=guide_legend(nrow = 3)) +
  scale_linetype_manual(name='', breaks=l, values=line_type,
                        labels=lab_strata[c(1,3,2,4,6,5)],
                        guide=guide_legend(nrow = 3)) +
  theme(legend.text=element_markdown(),
        plot.title = element_markdown())
```

It seems that the only susceptible flies are the _ken^1^_
